---
title: "Analysis of gene classes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Variables for shell scripts are stored in the files "shell_variables_ChIP-seq.sh" and "shell_variables_RNA-seq.sh" and can be modified as needed:

```{bash variables}
cat shell_variables_ChIP-seq.sh
cat shell_variables_RNA-seq.sh
```

#Assign genes to classes based on RNAP occupancy and gene expression

Gene classification criteria based on rnajA KO/WT comparison:

* class I: >=120% RNAP occupancy (ChIP-seq) and significantly upregulated (RNA-seq)

* class II: 0-80% RNAP occupancy (ChIP-seq) and significantly downregulated (RNA-seq)

* class III: <120% RNAP occupancy in WT (ChIP-seq) and significantly upregulated (RNA-seq)

* class IV: >=120% RNAP occupancy (ChIP-seq), and no significant change or significantly downregulated (RNA-seq)

* class V: all other genes with complete ChIP-seq and RNA-seq data

```{r, eval=FALSE}
chip <- read.delim('./coverage_ChIP-seq/ChIP-seq_gene_coverage.txt', 
                   sep = '\t', header = TRUE, stringsAsFactors = FALSE)
chip <- cbind(chip, chip$cov_ko / chip$cov_wt)
colnames(chip)[8] <- 'cov_ratio'
rna <- read.csv('./DESeq2results_rnjAKO_vs_WT.txt', 
                header = TRUE, stringsAsFactors = FALSE, row.names = 1)
rna <- rna[!is.na(rna$log2FoldChange), ]
chip <- chip[chip$Name %in% rownames(rna), ]
rna <- rna[rownames(rna) %in% chip$Name, ]
chip <- chip[order(chip$Name), ]
rna <- rna[order(rownames(rna)), ]
which(chip$Name != rownames(rna))
data <- cbind(chip, rna[, c('log2FoldChange', 'padj')])

padj.threshold <- 0.05  # False dicovery rate
class_I <- data[which(data$cov_ratio >= 1.2 & 
                      data$padj <= padj.threshold & data$log2FoldChange > 0), ]
class_II <- data[which(data$cov_ratio <= 0.8 & 
                       data$padj <= padj.threshold & data$log2FoldChange < 0), ]
class_III <- data[which(data$cov_ratio < 1.2 & 
                        data$padj <= padj.threshold & data$log2FoldChange > 0), ]
class_IV <- data[which(data$cov_ratio >= 1.2 & 
                         ((data$padj <= padj.threshold & data$log2FoldChange < 0) |
                          data$padj > padj.threshold)), ]
class_V <- data[which(!data$Name %in% 
                        c(class_I$Name, class_II$Name, class_III$Name, class_IV$Name)), ]

#pdf('./images/classes.pdf')
x.lim <- range(data$cov_ratio, na.rm = TRUE, finite = TRUE)
y.lim <- range(data$log2FoldChange, na.rm = TRUE, finite = TRUE)
plot(data[data$Name %in% class_I$Name, 'cov_ratio'], 
     data[data$Name %in% class_I$Name, 'log2FoldChange'],
     log = 'x', xlim = x.lim, ylim = y.lim, 
     xlab = 'RNAP occupancy', ylab = 'mRNA expression')
points(data[data$Name %in% class_II$Name, 'cov_ratio'], 
       data[data$Name %in% class_II$Name, 'log2FoldChange'], 
       col = 'red')
points(data[data$Name %in% class_III$Name, 'cov_ratio'], 
       data[data$Name %in% class_III$Name, 'log2FoldChange'], 
       col = 'blue')
points(data[data$Name %in% class_IV$Name, 'cov_ratio'], 
       data[data$Name %in% class_IV$Name, 'log2FoldChange'], 
       col = 'green')
legend(x = 'bottomright', legend = c('I', 'II', 'III', 'IV'), 
       fill = c('black', 'red', 'blue', 'green'))
#dev.off()

# save class gene lists
write.table(class_I[, c('Name', 'locus_tag', 'cov_wt', 'cov_ko', 'cov_ratio',
                        'log2FoldChange', 'padj')], 
            file = 'class_I.txt', quote = FALSE, sep = '\t', row.names = FALSE)
write.table(class_II[, c('Name', 'locus_tag', 'cov_wt', 'cov_ko', 'cov_ratio',
                        'log2FoldChange', 'padj')], 
            file = 'class_II.txt', quote = FALSE, sep = '\t', row.names = FALSE)
write.table(class_III[, c('Name', 'locus_tag', 'cov_wt', 'cov_ko', 'cov_ratio',
                        'log2FoldChange', 'padj')], 
            file = 'class_III.txt', quote = FALSE, sep = '\t', row.names = FALSE)
write.table(class_IV[, c('Name', 'locus_tag', 'cov_wt', 'cov_ko', 'cov_ratio',
                        'log2FoldChange', 'padj')], 
            file = 'class_IV.txt', quote = FALSE, sep = '\t', row.names = FALSE)
write.table(class_V[, c('Name', 'locus_tag', 'cov_wt', 'cov_ko', 'cov_ratio',
                        'log2FoldChange', 'padj')], 
            file = 'class_V.txt', quote = FALSE, sep = '\t', row.names = FALSE)

# produce class BED files
write.table(cbind(class_I$seqname, class_I$start - 1, class_I$end),
            file = './genome/class_I.bed', 
            quote = FALSE, sep = '\t', row.names = FALSE, col.names = FALSE)
write.table(cbind(class_II$seqname, class_II$start - 1, class_II$end),
            file = './genome/class_II.bed', 
            quote = FALSE, sep = '\t', row.names = FALSE, col.names = FALSE)
write.table(cbind(class_III$seqname, class_III$start - 1, class_III$end),
            file = './genome/class_III.bed', 
            quote = FALSE, sep = '\t', row.names = FALSE, col.names = FALSE)
write.table(cbind(class_IV$seqname, class_IV$start - 1, class_IV$end),
            file = './genome/class_IV.bed', 
            quote = FALSE, sep = '\t', row.names = FALSE, col.names = FALSE)
write.table(cbind(class_V$seqname, class_V$start - 1, class_V$end),
            file = './genome/class_V.bed', 
            quote = FALSE, sep = '\t', row.names = FALSE, col.names = FALSE)
```

#Create heatmaps and metagene profiles of RNAP occupancy for each class

The `bedGraphToBigWig` tool was downloaded from http://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/
on 2nd September 2019

```{bash}
deeptools --version
```

```{bash, eval=FALSE}
source shell_variables_ChIP-seq.sh

bedGraphToBigWig "${coverage_dir}wt.input-normalized.mean.bedgraph" "${genome_dir}chrom.sizes" "${coverage_dir}wt.input-normalized.mean.bw"
bedGraphToBigWig "${coverage_dir}KO.input-normalized.mean.bedgraph" "${genome_dir}chrom.sizes" "${coverage_dir}KO.input-normalized.mean.bw"

computeMatrix scale-regions -S "${coverage_dir}wt.input-normalized.mean.bw" "${coverage_dir}KO.input-normalized.mean.bw" -R "${genome_dir}class_I.bed" "${genome_dir}class_II.bed" "${genome_dir}class_III.bed" "${genome_dir}class_IV.bed" "${genome_dir}class_V.bed" --beforeRegionStartLength 200 --regionBodyLength 1000 --afterRegionStartLength 200 --outFileName "${coverage_dir}class_heatmap_matrix_ChIP-seq.mat.gz" --outFileNameMatrix "${coverage_dir}class_heatmap_matrix_ChIP-seq.values.tab" --numberOfProcessors "${CPU}"

plotHeatmap -m "${coverage_dir}class_heatmap_matrix_ChIP-seq.mat.gz" -out "${image_dir}class_heatmap_ChIP-seq.svg" --colorMap Greys --zMax 2 --averageTypeSummaryPlot mean --labelRotation 45 --startLabel "ATG" --endLabel "stop" --regionsLabel "class I genes" "class II genes" "class III genes" "class IV genes" "class V genes" --samplesLabel "wt" "rnjA KO" --whatToShow "heatmap and colorbar" --outFileNameMatrix "${coverage_dir}class_heatmap_ChIP-seq.tab"

plotProfile -m "${coverage_dir}class_heatmap_matrix_ChIP-seq.mat.gz" -out "${image_dir}class_profile_ChIP-seq.svg" --averageType mean --labelRotation 45 --startLabel "ATG" --endLabel "stop" --regionsLabel "class I" "class II" "class III" "class IV" "class V" --samplesLabel "wt" "rnjA KO" --yAxisLabel "RNAP occupancy" --plotTitle "average gene profiles" --outFileNameData "${coverage_dir}class_profile_ChIP-seq.tab"
```

#Create heatmaps and metagene profiles of transcript abundance for each class

```{bash, eval=FALSE}
source shell_variables_RNA-seq.sh

bedGraphToBigWig "${coverage_dir}wt.mean.bedgraph" "${genome_dir}chrom.sizes" "${coverage_dir}wt.mean.bw"
bedGraphToBigWig "${coverage_dir}KO.mean.bedgraph" "${genome_dir}chrom.sizes" "${coverage_dir}KO.mean.bw"

computeMatrix scale-regions -S "${coverage_dir}wt.mean.bw" "${coverage_dir}KO.mean.bw" -R "${genome_dir}class_I.bed" "${genome_dir}class_II.bed" "${genome_dir}class_III.bed" "${genome_dir}class_IV.bed" "${genome_dir}class_V.bed" --beforeRegionStartLength 200 --regionBodyLength 1000 --afterRegionStartLength 200 --outFileName "${coverage_dir}class_heatmap_matrix_RNA-seq.mat.gz" --outFileNameMatrix "${coverage_dir}class_heatmap_matrix_RNA-seq.values.tab" --numberOfProcessors "${CPU}"

plotHeatmap -m "${coverage_dir}class_heatmap_matrix_RNA-seq.mat.gz" -out "${image_dir}class_heatmap_RNA-seq.svg" --colorMap Greys --zMax 20 --averageTypeSummaryPlot mean --labelRotation 45 --startLabel "ATG" --endLabel "stop" --regionsLabel "class I genes" "class II genes" "class III genes" "class IV genes" "class V genes" --samplesLabel "wt" "rnjA KO" --whatToShow "heatmap and colorbar" --outFileNameMatrix "${coverage_dir}class_heatmap_RNA-seq.tab"

plotProfile -m "${coverage_dir}class_heatmap_matrix_RNA-seq.mat.gz" -out "${image_dir}class_profile_RNA-seq.svg" --averageType mean --labelRotation 45 --startLabel "ATG" --endLabel "stop" --regionsLabel "class I" "class II" "class III" "class IV" "class V" --samplesLabel "wt" "rnjA KO" --yAxisLabel "transcript abundance" --plotTitle "average gene profiles" --outFileNameData "${coverage_dir}class_profile_RNA-seq.tab"
```

#Replot average profiles using R

```{r}
colors <- c('blue', '#ffb400', 'green', 'red', 'grey30', 'darkgrey')
lty = 3
profile.rna <- t(read.delim('./coverage_RNA-seq/class_profile_RNA-seq.tab', 
                            header = FALSE, skip = 2))
profile.chip <- t(read.delim('./coverage_ChIP-seq/class_profile_ChIP-seq.tab', 
                            header = FALSE, skip = 2))
#pdf('./images/class_profiles.log.pdf', width = 16, height = 10)
svg('./images/class_profiles.log.svg', width = 16, height = 10)
par(mfrow = c(1, 2), lwd = 3, cex = 1.6, las = 1)

# RNA-seq
# WT
ylim <- range(as.numeric(profile.rna[-c(1, 2), ]))
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
points(profile.rna[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.rna[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.rna[-c(1, 2), 4], type = 'l', col = colors[4])
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors[5])
# rnjA KO
points(profile.rna[-c(1, 2), 6], type = 'l', col = colors[1], lty = lty)
points(profile.rna[-c(1, 2), 7], type = 'l', col = colors[2], lty = lty)
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors[3], lty = lty)
points(profile.rna[-c(1, 2), 9], type = 'l', col = colors[4], lty = lty)
points(profile.rna[-c(1, 2), 10], type = 'l', col = colors[5], lty = lty)
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(1, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), col = colors[6])

# ChIP-seq
# WT
#ylim <- range(as.numeric(profile.chip[-c(1, 2), ]))
plot(profile.chip[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = c(0.174, 5), xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
points(profile.chip[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.chip[-c(1, 2), 4], type = 'l', col = colors[4])
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors[5])
# rnjA KO
points(profile.chip[-c(1, 2), 6], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 7], type = 'l', col = colors[2], lty = lty)
points(profile.chip[-c(1, 2), 8], type = 'l', col = colors[3], lty = lty)
points(profile.chip[-c(1, 2), 9], type = 'l', col = colors[4], lty = lty)
points(profile.chip[-c(1, 2), 10], type = 'l', col = colors[5], lty = lty)
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(0.133, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), col = colors[6])
legend(x = 1, y = 11, 
       legend = c('class I', 'class II', 'class III', 'class IV', 'class V', 'WT', 'rnjA KO'), 
       col = c(colors[1:5], 'black', 'black'), 
       lty = c(1, 1, 1, 1, 1, 1, lty), bty = 'n', ncol = 3, xpd = TRUE)
dev.off()


# Class V only
svg('./images/class_profiles.log.class_V.svg', width = 16, height = 10)
par(mfrow = c(1, 2), lwd = 3, cex = 1.6, las = 1)
# RNA-seq
# WT
ylim <- range(as.numeric(profile.rna[-c(1, 2), ]))
plot(profile.rna[-c(1, 2), 5], 
     type = 'l', log = 'y', col = colors[5], ylim = ylim, xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
# rnjA KO
points(profile.rna[-c(1, 2), 10], type = 'l', col = colors[5], lty = lty)
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(1, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), col = colors[6])

# ChIP-seq
# WT
#ylim <- range(as.numeric(profile.chip[-c(1, 2), ]))
plot(profile.chip[-c(1, 2), 5], 
     type = 'l', log = 'y', col = colors[5], ylim = c(0.174, 5), xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
# rnjA KO
points(profile.chip[-c(1, 2), 10], type = 'l', col = colors[5], lty = lty)
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(0.133, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), col = colors[6])
dev.off()
```

```{r, include=FALSE, eval=FALSE}
# different colour schemes
pdf('./images/class_profiles.log2.pdf', width = 16, height = 10)
par(mfrow = c(1, 2), lwd = 3, cex = 1.3)
colors.wt <- c('blue1', 'blue2', 'blue3', 'blue4', 'darkgrey')
colors.ko <- c('red1', 'red2', 'red3', 'red4')
# RNA-seq
# WT
ylim <- range(as.numeric(profile.rna[-c(1, 2), ]))
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors.wt[1], ylim = ylim, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'WT')
points(profile.rna[-c(1, 2), 2], type = 'l', col = colors.wt[2])
points(profile.rna[-c(1, 2), 3], type = 'l', col = colors.wt[3])
points(profile.rna[-c(1, 2), 4], type = 'l', col = colors.wt[4])
abline(v = c(20, 120), col = colors.wt[5])
# rnjA KO
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors.ko[1], lty = lty)
points(profile.rna[-c(1, 2), 6], type = 'l', col = colors.ko[2], lty = lty)
points(profile.rna[-c(1, 2), 7], type = 'l', col = colors.ko[3], lty = lty)
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors.ko[4], lty = lty)
abline(v = c(20, 120), col = colors.wt[5])

# ChIP-seq
# WT
ylim <- range(as.numeric(profile.chip[-c(1, 2), ]))
plot(profile.chip[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors.wt[1], ylim = ylim, xaxs = 'i',
     ylab = 'RNAP occupancy', xlab = '', main = 'WT')
points(profile.chip[-c(1, 2), 2], type = 'l', col = colors.wt[2])
points(profile.chip[-c(1, 2), 3], type = 'l', col = colors.wt[3])
points(profile.chip[-c(1, 2), 4], type = 'l', col = colors.wt[4])
abline(v = c(20, 120), col = colors.wt[5])
# rnjA KO
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors.ko[1], lty = lty)
points(profile.chip[-c(1, 2), 6], type = 'l', col = colors.ko[2], lty = lty)
points(profile.chip[-c(1, 2), 7], type = 'l', col = colors.ko[3], lty = lty)
points(profile.chip[-c(1, 2), 8], type = 'l', col = colors.ko[4], lty = lty)
abline(v = c(20, 120), col = colors.wt[5])


# grayscale colour schemes
par(mfrow = c(1, 2), lwd = 3, cex = 1.3)
colors <- c('gray0', 'gray25', 'gray50', 'gray75', 'darkgrey')
# RNA-seq
# WT
ylim <- range(as.numeric(profile.rna[-c(1, 2), ]))
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'WT')
points(profile.rna[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.rna[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.rna[-c(1, 2), 4], type = 'l', col = colors[4])
abline(v = c(20, 120), col = colors[5])
# rnjA KO
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors[1], lty = lty)
points(profile.rna[-c(1, 2), 6], type = 'l', col = colors[2], lty = lty)
points(profile.rna[-c(1, 2), 7], type = 'l', col = colors[3], lty = lty)
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors[4], lty = lty)
abline(v = c(20, 120), col = colors[5])

# ChIP-seq
# WT
ylim <- range(as.numeric(profile.chip[-c(1, 2), ]))
plot(profile.chip[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i',
     ylab = 'RNAP occupancy', xlab = '', main = 'WT')
points(profile.chip[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.chip[-c(1, 2), 4], type = 'l', col = colors[4])
abline(v = c(20, 120), col = colors[5])
# rnjA KO
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 6], type = 'l', col = colors[2], lty = lty)
points(profile.chip[-c(1, 2), 7], type = 'l', col = colors[3], lty = lty)
points(profile.chip[-c(1, 2), 8], type = 'l', col = colors[4], lty = lty)
abline(v = c(20, 120), col = colors[5])


# plotted by classes, RNA-seq and ChIP-seq in one plot
colors <- c('blue', 'red', 'darkgrey')
par(mfrow = c(2, 2), lwd = 3)
ylim <- range(c(as.numeric(profile.rna[-c(1, 2), ]), as.numeric(profile.chip[-c(1, 2), ])))
# class I
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', 
     ylab = 'normalized coverage', xlab = '', main = 'class I')
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 1], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors[2], lty = lty)
abline(v = c(20, 120), col = colors[5])
# class II
plot(profile.rna[-c(1, 2), 2], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i',
     ylab = 'normalized coverage', xlab = '', main = 'class II')
points(profile.rna[-c(1, 2), 6], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 2], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 6], type = 'l', col = colors[2], lty = lty)
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors[4], lty = lty)
abline(v = c(20, 120), col = colors[5])
# class III
plot(profile.rna[-c(1, 2), 3], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i',
     ylab = 'normalized coverage', xlab = '', main = 'class III')
points(profile.rna[-c(1, 2), 7], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 3], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 7], type = 'l', col = colors[2], lty = lty)
abline(v = c(20, 120), col = colors[5])
# class IV
plot(profile.rna[-c(1, 2), 4], 
      type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', 
      ylab = 'normalized coverage', xlab = '', main = 'class IV')
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 4], type = 'l', col = colors[1], lty = lty)
points(profile.chip[-c(1, 2), 8], type = 'l', col = colors[2], lty = lty)
abline(v = c(20, 120), col = colors[5])
dev.off()


# plotted by classes, RNA-seq and ChIP-seq separately
pdf('./images/class_profiles.log3.pdf', width = 16, height = 20)
colors <- c('blue', 'red', 'darkgrey')
par(mfrow = c(4, 2), lwd = 3, cex = 1.3)
ylim.rna <- range(as.numeric(profile.rna[-c(1, 2), ]))
ylim.chip <- range(as.numeric(profile.chip[-c(1, 2), ]))
# class I
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim.rna, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'class I')
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors[2])
plot(profile.chip[-c(1, 2), 1],  
     type = 'l', log = 'y', col = colors[1], ylim = ylim.chip, xaxs = 'i', 
     ylab = 'RNAP occupancy', xlab = '', main = 'class I')
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors[2])
abline(v = c(20, 120), col = colors[5])
# class II
plot(profile.rna[-c(1, 2), 2], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim.rna, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'class II')
points(profile.rna[-c(1, 2), 6], type = 'l', col = colors[2])
plot(profile.chip[-c(1, 2), 2],  
     type = 'l', log = 'y', col = colors[1], ylim = ylim.chip, xaxs = 'i', 
     ylab = 'RNAP occupancy', xlab = '', main = 'class II')
points(profile.chip[-c(1, 2), 6], type = 'l', col = colors[2])
abline(v = c(20, 120), col = colors[5])
# class III
plot(profile.rna[-c(1, 2), 3], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim.rna, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'class II')
points(profile.rna[-c(1, 2), 7], type = 'l', col = colors[2])
plot(profile.chip[-c(1, 2), 3],  
     type = 'l', log = 'y', col = colors[1], ylim = ylim.chip, xaxs = 'i', 
     ylab = 'RNAP occupancy', xlab = '', main = 'class III')
points(profile.chip[-c(1, 2), 7], type = 'l', col = colors[2])
abline(v = c(20, 120), col = colors[5])
# class IV
plot(profile.rna[-c(1, 2), 4], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim.rna, xaxs = 'i', 
     ylab = 'transcript abundance', xlab = '', main = 'class II')
points(profile.rna[-c(1, 2), 8], type = 'l', col = colors[2])
plot(profile.chip[-c(1, 2), 4],  
     type = 'l', log = 'y', col = colors[1], ylim = ylim.chip, xaxs = 'i', 
     ylab = 'RNAP occupancy', xlab = '', main = 'class IV')
points(profile.chip[-c(1, 2), 8], type = 'l', col = colors[2])
abline(v = c(20, 120), col = colors[5])
dev.off()
```

#Relative average gene profiles

```{bash, eval=TRUE}
# ChIP-seq
source shell_variables_ChIP-seq.sh

bigwigCompare -b1 "${coverage_dir}KO.input-normalized.mean.bw" -b2 "${coverage_dir}wt.input-normalized.mean.bw" --operation ratio --binSize 1 --numberOfProcessors "${CPU}" --outFileName "${coverage_dir}ratio.input-normalized.mean.bw"

computeMatrix scale-regions -S "${coverage_dir}ratio.input-normalized.mean.bw" -R "${genome_dir}class_I.bed" "${genome_dir}class_II.bed" "${genome_dir}class_III.bed" "${genome_dir}class_IV.bed" "${genome_dir}class_V.bed" --beforeRegionStartLength 200 --regionBodyLength 1000 --afterRegionStartLength 200 --outFileName "${coverage_dir}class_heatmap_matrix_ChIP-seq.ratio.mat.gz" --numberOfProcessors "${CPU}"

plotProfile -m "${coverage_dir}class_heatmap_matrix_ChIP-seq.ratio.mat.gz" -out "${image_dir}class_profile_ChIP-seq.ratio.svg" --averageType mean --labelRotation 45 --startLabel "START" --endLabel "STOP" --regionsLabel "class I" "class II" "class III" "class IV" "class V" --samplesLabel "rnjA KO vs WT" --yAxisLabel "relative RNAP occupancy" --plotTitle "average gene profiles" --outFileNameData "${coverage_dir}class_profile_ChIP-seq.ratio.tab"

# RNA-seq
source shell_variables_RNA-seq.sh

bigwigCompare -b1 "${coverage_dir}KO.mean.bw" -b2 "${coverage_dir}wt.mean.bw" --operation ratio --binSize 1 --numberOfProcessors "${CPU}" --outFileName "${coverage_dir}ratio.mean.bw"

computeMatrix scale-regions -S "${coverage_dir}ratio.mean.bw" -R "${genome_dir}class_I.bed" "${genome_dir}class_II.bed" "${genome_dir}class_III.bed" "${genome_dir}class_IV.bed" "${genome_dir}class_V.bed" --beforeRegionStartLength 200 --regionBodyLength 1000 --afterRegionStartLength 200 --outFileName "${coverage_dir}class_heatmap_matrix_RNA-seq.ratio.mat.gz" --numberOfProcessors "${CPU}"

plotProfile -m "${coverage_dir}class_heatmap_matrix_RNA-seq.ratio.mat.gz" -out "${image_dir}class_profile_RNA-seq.ratio.svg" --averageType mean --labelRotation 45 --startLabel "START" --endLabel "STOP" --regionsLabel "class I" "class II" "class III" "class IV" "class V" --samplesLabel "rnjA KO vs WT" --yAxisLabel "relative transcript abundance" --plotTitle "average gene profiles" --outFileNameData "${coverage_dir}class_profile_RNA-seq.ratio.tab"
```

```{r}
class.lengths <- c(nrow(class_I), nrow(class_II), nrow(class_III), nrow(class_IV), nrow(class_V))
colors <- c('blue', '#ffb400', 'green', 'red', 'grey30', 'lightgrey')
profile.rna <- t(read.delim('./coverage_RNA-seq/class_profile_RNA-seq.ratio.tab', 
                            header = FALSE, skip = 2))
profile.chip <- t(read.delim('./coverage_ChIP-seq/class_profile_ChIP-seq.ratio.tab', 
                            header = FALSE, skip = 2))
# error intervals for average profiles
chip.matrix <- read.delim('./coverage_ChIP-seq/class_heatmap_matrix_ChIP-seq.ratio.mat.gz', 
                          header = FALSE, skip = 1)
chip.matrix <- as.matrix(chip.matrix[, -c(1:6)])
rna.matrix <- read.delim('./coverage_RNA-seq/class_heatmap_matrix_RNA-seq.ratio.mat.gz', 
                         header = FALSE, skip = 1)
rna.matrix <- as.matrix(rna.matrix[, -c(1:6)])
chip.spread <- matrix(data = NA, nrow = 140, ncol = 5)
rna.spread <- matrix(data = NA, nrow = 140, ncol = 5)
for (i in 1:ncol(chip.spread)){
  for (j in 1:nrow(chip.spread)){
    chip.spread[j, i] <- sd(chip.matrix[sum(class.lengths[0:(i-1)]):sum(class.lengths[0:i]), j],
                            na.rm = TRUE)
    rna.spread[j, i] <- sd(rna.matrix[sum(class.lengths[0:(i-1)]):sum(class.lengths[0:i]), j],
                           na.rm = TRUE)
  }
}

#pdf('./images/class_profiles.ratio.log.pdf', width = 16, height = 10)
svg('./images/class_profiles.ratio.log.svg', width = 16, height = 10)
par(mfrow = c(1, 2), lwd = 3, cex = 1.6, las = 1)
colors.alpha <- c('#0000ff50', '#ffb40050', '#00ff0050', '#ff000050', '#30303050')
# RNA-seq
ylim <- range(as.numeric(profile.rna[-c(1, 2), ]))
plot(profile.rna[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
points(profile.rna[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.rna[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.rna[-c(1, 2), 4], type = 'l', col = colors[4])
points(profile.rna[-c(1, 2), 5], type = 'l', col = colors[5])
# rect(1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 1]), 
#      1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 1]) + rna.spread[, 1], 
#      lwd = 4, border = colors.alpha[1])
# rect(1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 2]), 
#      1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 2]) + rna.spread[, 2], 
#      lwd = 4, border = colors.alpha[2])
# rect(1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 3]), 
#      1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 3]) + rna.spread[, 3], 
#      lwd = 4, border = colors.alpha[3])
# rect(1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 4]), 
#      1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 4]) + rna.spread[, 4], 
#      lwd = 4, border = colors.alpha[4])
# rect(1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 5]), 
#      1:nrow(profile.rna), as.numeric(profile.rna[-c(1, 2), 5]) + rna.spread[, 5], 
#      lwd = 4, border = colors.alpha[5])
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(0.3, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), h = 1, col = colors[6])

# ChIP-seq
ylim <- range(as.numeric(profile.chip[-c(1, 2), ]))
plot(profile.chip[-c(1, 2), 1], 
     type = 'l', log = 'y', col = colors[1], ylim = ylim, xaxs = 'i', xaxt = 'n',
     ylab = '', xlab = '', main = '')
points(profile.chip[-c(1, 2), 2], type = 'l', col = colors[2])
points(profile.chip[-c(1, 2), 3], type = 'l', col = colors[3])
points(profile.chip[-c(1, 2), 4], type = 'l', col = colors[4])
points(profile.chip[-c(1, 2), 5], type = 'l', col = colors[5])
# rect(1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 1]), 
#      1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 1]) + chip.spread[, 1], 
#      lwd = 4, border = colors.alpha[1])
# rect(1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 2]), 
#      1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 2]) + chip.spread[, 2], 
#      lwd = 4, border = colors.alpha[2])
# rect(1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 3]), 
#      1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 3]) + chip.spread[, 3], 
#      lwd = 4, border = colors.alpha[3])
# rect(1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 4]), 
#      1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 4]) + chip.spread[, 4], 
#      lwd = 4, border = colors.alpha[4])
# rect(1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 5]), 
#      1:nrow(profile.chip), as.numeric(profile.chip[-c(1, 2), 5]) + chip.spread[, 5], 
#      lwd = 4, border = colors.alpha[5])
axis(side = 1, at = c(1, 20, 120, 140), lwd.ticks = 2, labels = FALSE)
text(x = c(1, 20, 120, 140), y = rep(0.75, 4),
     labels = c('-0.2 kb', 'START', 'STOP', '0.2 kb'), srt = 45, xpd = TRUE, pos = 1)
abline(v = c(20, 120), h = 1, col = colors[6])
legend(x = 1, y = 1.2, lty = 1,
       legend = c('class I', 'class II', 'class III', 'class IV', 'class V'), 
       col = colors[1:5], bty = 'n', ncol = 2, xpd = TRUE)
dev.off()
```

***

```{r}
sessionInfo()
```